---
title: "Ooh na na... where are my sodium labs?"
description: |
  The NA bug, or, what happens when the same word is used in different contexts. [5 min read]
author:
  - name: Chloe Pou-Prom
    affiliation: DSAA, Unity Health Toronto
    affiliation_url: https://chartdatascience.ca
date: 2022-05-02
output:
  distill::distill_article:
    toc: true
categories:
  - language-R
  - project-chartwatch
  - post-miscellaneous
draft: true
preview: preview.PNG
bibliography: references.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load_date, echo = FALSE}
library(here)
library(xts)
library(dygraphs)
library(dplyr)
library(ggplot2)
data <- read.csv(file.path(here::here(), "_posts/2022-05-02-the-na-bug", "daily_electrolyte_counts.csv"), stringsAsFactors = FALSE) %>%
  mutate(date = as.Date(date))

daily_counts_pre_fix <- data %>%
    filter(cohort == "pre-fix")

daily_counts_post_fix <- data %>%
    filter(cohort == "post-fix")
```

## Silent deployment

Our team had been working endlessly on developing [CHARTwatch](https://www.chartdatascience.ca/the-projects/an-early-warning-system-for-general-internal-medicine-), an early warning system and in November 2019, we were ready to move to our **silent deployment** phase. This phase involved deploying our system and, as the name suggests it, running it _silently_. More concretely, our entire pipeline was running (from data extraction to data processing to model prediction), but no outputs were going to the end-user. 

Typically, the goal of the silent deployment phase is to uncover unexpected behaviors with the data, system, or model. During model development and evaluation, we had only worked with _historical extracts of the data_. When moving from historical data to _live data_, there's the risk of running into data issues [@cohen_problems_2021].

- The data can be different due to **external factors**. For example, all of our models were trained on data prior to COVID-19, but shortly after the beginning of our silent deployment phase, we began to observe cases of COVID-19 in the hospital.

- The data can be different due to **data entry errors**. For example, a body temperature could incorrectly be entered as 3700 °C instead of 37.00 °C.

- The data can be different due to **selection bias**. For example, during training we excluded patients with really short and really long visits, as they were rare. However, we may encounter these kinds of visits in the live data.


## Monitoring labs

We had set up a monitoring dashboard to measure model inputs and model outputs. On close inspection, we made a discovery that was unquestioNAbly odd... no sodium labs had been measured since we had moved to silent testing!

```{r missingna, fig.cap="Daily counts of lab measurements: this includes counts for calcium (CA), chloride (CL), glucose (GLPOC), potassium (K), and sodium (NA)."}
daily_counts_pre_fix %>%
    select(-date, -cohort) %>%
    xts(order.by = daily_counts_pre_fix$date) %>%
    dygraph(main = "Electrolyte counts")  %>%
    dyOptions(stackedGraph = TRUE) %>%
    dyLegend(width = 400)
```

This made NA sense at all. Sodium is measured in routinely ordered blood tests. It'll usually get ordered alongside other tests (such as calcium, chloride, glucose, and potassium) as part of a [basic metabolic panel](https://www.healthline.com/health/sodium-blood). In Figure \@ref(fig:missingna), we look at the daily counts of labs on units in which CHARTwatch was silently deployed. 
The other labs were regularly measured, but our pipeline had not detected a single sodium lab. There was NA way sodium would be missing!

## The NA bug

After hours of detective work, we found the issue:

- In R, the programming language we used to develop CHARTwatch, the symbol `NA` stands for "not available" and is used to represent missing data.

- In chemistry, `Na` is the symbol used to represent the chemical element of sodium.

```{r fixedna, fig.cap="Daily counts of lab measurements after fixing the NA bug"}
data %>%
    select(-date, -cohort) %>%
    xts(order.by = data$date) %>%
    dygraph(main = "Electrolyte counts")  %>%
    dyOptions(stackedGraph = TRUE) %>%
    dyLegend(width = 400)
```


Depending on the context, the symbol meant something different! Our data extraction pipeline was interpreting the chemical element `Na` as "not available"!

We were able to fix the _NA bug_ and after re-deploying, sodium counts were back to normal (as seen in Figure \@ref(fig:fixedna)).

The fix was quite simple, but the problem we uncovered lead to plenty of follow-up questions! 

- Were there other cases where the same symbol meant two different things based on the context? 

- What does our electronic health record use to represent a missing value? Do they go with a number that's biologically impossible? (e.g., a body temperature of -1000) Do they use a specific symbol/term? (e.g., "not measured", "missing") 

- How are these decisions made? 

Recently, there's been a push for improvement in data quality standards, such as "Datasheets for Datasets" [@gebru_datasheets_2021] and the explosion of [features stores, model stores, and evaluation stores](https://www.mihaileric.com/posts/mlops-is-a-mess/)^[What kind of "store" do we think is next? 🤔].


## Takeaways

- NA (sodium) ≠ NA (not available)

- Silent deployment is important.

- Thorough metadata and data quality standards are important to mitigating these kinds of issues.